---
title: "Introduction to Data Science"
# subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
subtitle: "Lecture 2: Data Structures and Sources"
author: "Luka Sikic"
date: "Catholic University of Croatia | [GH](https://github.com/lusiki/DataScience)" #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{css, echo=FALSE}
# CSS for including pauses in printed PDF output (see bottom of lecture)
@media print {
  .has-continuation {
    display: block !important;
  }
}
```

```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.

options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
```


# Table of contents
<br><br><br>

1. [Structured and unstructured data](#structure)

2. [Some important types of data](#dpractice)

3. [Big Data in Social Sciences](#BData)

4. [Sources](#sources)



---
class: inverse, center, middle
name: structure

# Structured and unstructured data

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Structured Data

<br><br><br><br>
- observations with characteristics, mostly organized in a table (rows and columns)
<br>
- scientifically collected data, telephone directory
<br>
- a smaller part of the data overall

---

# Unstructured Data

<br><br><br><br>
- data without a standard organizational hierarchy
<br>
- Facebook posts, Twitter, server logs, genetic sequence of nucleotides, text data
<br>
- probably more than 80% of all data
<br>
- require adjustment before analysis

---

# Quantitative data

<br><br>
- numbers
<br>
- mathematical procedures
<br>
- average
<br>
- time trend
<br>
- threshold effects

---

# Qualitative data
<br><br>
- "natural" categories
<br>
- language
<br>
- the most frequent observation, unique values


---
class: inverse, center, middle
name: dpractice

# Some important types of data

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

## Big Data



- due to their size, they cannot be stored on standard relational databases
<br>
- unstructured, semi-structured and structured data
<br>
- from terabytes to zettabytes in size
<br>
- basis for machine learning, AI, predicting the future

--- 

## Time series

- each observation has a time stamp (event or processing)
<br>
- eg. behavioral data (web) used to predict action models
<br>
- eg. financial trading

---


## Machine data

- system produced data
<br>
- eg PC logs, mobile phones, smart cars, applications, API, cloud applications, sensor signals
<br>
- especially useful for threat identification, process quality control, forecasting

---

## Spatial-temporal data

- geographical and temporal stamp
<br>
- demonstrate locational dynamics in time
<br>
- tracking of vehicles, changes in the population, detection of anomalies in the (e.g. telecom) network

---

## Open data

- available to the public according to the "rules for use"
<br>
- currently mostly limited to scientific use

---

## Black data

- unused data
<br>
- mostly institutionally owned (companies, institutes, think-tanks, faculties)

---

## Data in real time

- high demand and "hot trend"
<br>
- 5G
<br>
- medicine, transport, trade
<br>
- great potential

---

## Genomic data

- DNA analysis in order to identify the effects of drugs and treat diseases
<br>
- analytics is the main limitation
<br>
- collection and processing of data is difficult

---

## Operational data

- insights into processes beyond analysts and specialists
<br>
- data on internships
<br>
- useful for reverse engineering

---

## High-dimensional data

- popularized through IT support for facial recognition
<br>
- due to the complexity of the transformations, they require new ways of recording and manipulation


---
class: inverse, center, middle
name: BData

# Big Data in Social Sciences

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Characteristics 
<br><br>
- high volume, scope, speed and variety
<br>
- geographic, textual, and social media data
<br>
- availability in real time, greater scope, new indicators and smaller structures
<br>
- in the business sector and natural sciences, BD has already become a standard
<br>
- social sciences and public institutions (governments) are late in accepting BD

---

# Potential
<br><br>
- new insights into society, politics and economy
<br>
- new research questions
<br>
- more accurate predictions
<br>
- better public policies

---

# Challenges
<br><br>
- access to data
<br>
- data management and analysis implementation (T.Khun!?)
<br>
- asking the right questions

---
class: inverse, center, middle
name: sources

# Data Sources

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---


# Overview

<br>
<br>
<br>

.pull-left[

#### **Classic**

- An experiment
- Examination
- An interview
- Structured databases

]

.pull-right[

#### **Modern (BigData)**

- Web
- Social networks
- Applications
- Sensors
- Modern data management

]


.footnote[[*]Machine retrieval of (modern) data sources!]

---

# Old school sources

### Experiment

* High degree of control
* Planned design
* Longitudinal and case studies
* The problem with generalization 
]

.pull-right[
### Examination

* Target population
* Larger sample
* Analysis of behavior, feelings, experiences
* The problem of representativeness of the sample
* Custom design required
]

---

# Old school sources

.pull-left[

### Interview

* Profiled pattern
* Smaller sample
* Suitable for in-depth analysis
* Depends on the researcher
* Difficulty reproducing results
]


.pull-right[

### Structured dta bases

* Enable quantitative analysis
* Macro and mezzo level of analysis
* It is necessary to adjust the data
* Institutional and regulated data
* Machine download of data
]

---

# Classic

### Example of using a structured database ([gapminder](https://www.gapminder.org/))

```{r}
# učitaj paket
library(gapminder)
# pregledaj podatke
head(gapminder,8)
```

---

# Classic

### Example of using a structured database ([gapminder](https://www.gapminder.org/))

```{r eval=FALSE}
# učitaj pakete
library(gganimate)
library(gifski)
library(ggplot2)
# Izradi animiranu vizualizaciju
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) 
```

---

# Classic

### Example of using a structured database ([gapminder](https://www.gapminder.org/))

```{r echo=F}
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent)
```

---

# Modern data sources (BigData)


<br>
<br>
<br>
1. Web
<br>
<br>
2. Social networks
<br>
<br>
3. Applications
<br>
<br>
4. Sensory devices


---

# Modern data sources (BigData)


**Characteristics**:

1. Big
2. Fast
3. Various
4. Valuable
5. Legally regulated


<br><br>

* Internet, smart devices, smart sensors
* New data paradigm
* Advanced technological infrastructure


---

# Modern data sources (BigData)

### NEW RESEARCH PARADIGM

* Analysis starts with data (theory is secondary)
* Technical skills required
* Complex and memory (RAM) demanding analytics
* Novelties in the presentation and communication of results
* Business applications lead the way
* Application in social research is in its infancy (*computational social science*)
* It is necessary to define a new research architecture (ecosystem)

.footnote[[*] [Khun's revolution!?](https://journals.sagepub.com/doi/10.1177/2053951714528481)]


---

# Web


<br>
<br>
<br>
<br>
<br>
**Source of data for:**
1. explanation
<br>
<br>
2. modeling
<br>
<br>
3. "nowcasting"
<br>
<br>
4. prediction

---
# Web

<br>
<br>
<br>
<br>
**Data collection:**
<br>
<br>
1. [Web-Scraping](https://www.datacamp.com/community/tutorials/r-web-scraping-rvest)
<br>
<br>
2. API (*application programming interface*)

---
# Web
### Web scraping

```{r warning=FALSE}
# paket
library(rvest) 
url <- "https://en.wikipedia.org/wiki/List_of_Formula_One_drivers" 
page <- read_html(url)
drivers_F1 <- html_element(page, "table.sortable") %>%
  html_table()
head(drivers_F1)

```

---
# Web
### Web scraping
```{r }
str(drivers_F1)
```

---
# Web
### Web scraping
```{r}
drivers_F1 <- drivers_F1[-nrow(drivers_F1), ] # remove last row
drivers_F1$`Drivers' Championships` <- substr(drivers_F1$`Drivers' Championships`,
  start = 1, stop = 1
)

drivers_F1 %>%
  group_by(Nationality) %>%
  summarise(championship_country = sum(as.double(`Drivers' Championships`))) %>%
  arrange(desc(championship_country))
```

```{r}
drivers_F1 %>%
  group_by(`Driver name`) %>%
  summarise(championship_pilot = sum(as.double(`Drivers' Championships`))) %>%
  arrange(desc(championship_pilot))
```

```{r}
drivers_F1 %>%
  filter(`Pole positions` > 1) %>%
  ggplot(aes(x = as.double(`Pole positions`), y = as.double(`Drivers' Championships`))) +
  geom_point(position = "jitter") +
  labs(y = "Championships won", x = "Pole positions") +
  theme_minimal()
```





---
# Internet 
### Internet - Google trends

```{r echo=T}
# paketi
library (gtrendsR) 
library(ggplot2)
#kljucne rijeci
keywords <- c("zavjera", "cjepljenje", "prosvjedi")
#GEO
country <- c('HR')
#vremenski raspon
time <- paste("2018-01-01", Sys.Date())
#kanali
channel <- 'web'
# funkcija iz paketa
trends <- gtrends(keywords, gprop = channel,geo = country, time = time )
#samo 'interes kroz vrijeme'
time_trend <- trends$interest_over_time
```

---

# Internet 
### Internet - Google trends
 <br>
 
```{r echo = TRUE}
# paket
library(tidyverse)
# pregledaj podatke
tail(sample_n(time_trend,7),7)
```


---

# Internet 
### Internet - Google trends
 <br>
 
```{r echo=TRUE, fig.width=5, fig.height=3}
# paket
library(lubridate)
# napravi grafikon 
google_pretraga <- ggplot(data = time_trend,
               aes(x = date, y = hits, group = keyword, col = keyword)) +
               geom_line() +
               xlab('Vrijeme') + 
               ylab('Relativni interes') +
               theme_bw() +
               theme(legend.title = element_blank(),
                     legend.position = "bottom",
                     legend.text = element_text (size = 12)) + 
               ggtitle("Google volumen pretrage")
  
```

---

# Internet 
### Internet - Google trends

```{r  echo=TRUE, fig.width=4, fig.height=2}
# grafički prikaz
google_pretraga
```

---

# Internet 
### Internet - Google trends

```{r echo=TRUE, fig.width=5, fig.height=2}
# sve u jednom redu koda + "mjere"   :-) 
plot(gtrendsR::gtrends(keyword = c("zavjera", "cjepljenje", "prosvjedi","mjere"),
                       geo = "HR",
                       time = time))
```



### Examples of research based on Google trends:

1. [Home Sales Analysis](https://rpubs.com/tonytusharjr/newhomes)
<br>
<br>
2. [Demand for cars](https://aisel.aisnet.org/icis2017/DataScience/Presentations/1/)
<br>
<br>
3. [Tourism demand](https://www.researchgate.net/publication/336516335_Forecasting_international_tourism_demand_in_Croatia_using_Google_Trends)
<br>
<br>
4. [Unemployment](https://www.sciencedirect.com/science/article/abs/pii/S1544612319301072)
<br>
<br>
5. [Movements in stock markets](https://www.emerald.com/insight/content/doi/10.1108/JED-07-2019-0017/full/pdf?title=stock-market-activity-and-google -trends-the-case-of-a-developing-economy)
<br>
<br>
6. [Purchasing cinema tickets](https://www.tandfonline.com/doi/abs/10.1080/13504851.2011.613744)
<br>
<br>
7. [Political radicalism in China](https://martinctc.github.io/blog/vignette-google-trends-with-gtrendsr/)
<br>
<br>
8. [Political views](https://journals.sagepub.com/doi/full/10.1177/2378023118760414)

---
# API
### API - COVID19 data

```{r ,fig.width=6, fig.height=3, warning=FALSE}
# učitaj pakete
#remotes::install_github("joachim-gassen/tidycovid19")
library(tidycovid19)
library(zoo)
# preuzmi podatke
df <- download_merged_data(cached = TRUE, silent = TRUE)
# uredi podatke i napravi grafikon
covid <- df %>%
  filter(iso3c == "HRV") %>%
  mutate(
    new_cases = confirmed - lag(confirmed),
    ave_new_cases = rollmean(new_cases, 7, na.pad=TRUE, align="right")) %>%
  filter(!is.na(new_cases), !is.na(ave_new_cases)) %>%
  ggplot(aes(x = date)) +
  geom_bar(aes(y = new_cases), stat = "identity", fill = "lightblue") +
  geom_line(aes(y = ave_new_cases), color ="red") +
  theme_minimal() +
  xlab("Novi slučajevi") + 
  ylab("Datum") +
  ggtitle("COVID19 incidencija u Hrvatskoj")
```

---
# API
### API - COVID19 data


```{r, fig.width=6, fig.height=3}
# grafički prikaz
covid
```

--- 

# Social Media

<br>
**Popular social networks:**
<br>
<br>
- [Facebook](https://www.kaggle.com/chrisbow/an-introduction-to-facebook-ad-analysis-using-r) [(See also!)](https://www.residualthoughts.com /2018/02/12/analyzing-facebook-messages-in-r/)
<br>
- [LinkedIn](https://github.com/mpiccirilli/Rlinkedin)
<br>
- [Youtube](https://rpubs.com/arafath/YouTubeVideos_Analysis)
<br>
- [Instagram](https://github.com/JonasSchroeder/InstaCrawlR)
<br>
- TikTok
<br>
- Google+
<br>
- Tumblr
<br>
- Flickr
<br>
<br>
<br>
.footnote[[*] [Digital 2021:Croatia](https://datareportal.com/reports/digital-2021-croatia?rq=Croatia)]

---

# Social Media

<br>
<br>
<br>
### The most important trends

<br>
- Business applications lead the way
<br>
- Perspective research field in political science, communication science, economics
<br>
- [API-calypse](https://snurb.info/files/2019/After%20the%20%27APIcalypse%27.pdf) and the *Cambridge Analytica* case; [Social Dilemma](https://www.thesocialdilemma.com/)
<br>
- What to do next!?
<br>
<br>
<br>

.footnote[[*] [College!](https://lusiki.github.io/Learning-Social-Media-Analytics/)]
<br>



---
# Social Media
<br>
#### Croatian influencers on Facebook
<br>
<br>
- data downloaded from the [Apify](https://apify.com/) service
<br>
- data with payment
<br>
- FB data available only for public profiles
<br>
- technological brands in Croatia
<br>
- about a month
 
---
# Social Media
#### Croatian influencers on Facebook (*impressions*)
<br>

<br>
```{r importFacebook, echo=FALSE, message=FALSE, warning=FALSE}

library(readxl)
library(janitor)


site_data_columns <- read_excel("./02-data/09-21.xlsx",
                               sheet = "Facebook")

site_data_column_names <- colnames(site_data_columns)
column_data_types <- c()
for(i in 1 : length(site_data_column_names)){  

    #where date is a column name
    if(site_data_column_names[i] == "date"){
         column_data_types[i] <- "date"

         #where result is a column name
         } else if (site_data_column_names[i] == "result") {
                      column_data_types[i] <- "numeric"

         } else{
                column_data_types[i] <- "text"
        }
}
column_data_types[5] <- "date"



facebook <- readxl::read_excel("./02-data/09-21.xlsx",
                               sheet = "Facebook",
                               col_types = column_data_types) %>% 
  row_to_names(.,1) %>% 
  clean_names() %>%
  select(-indexed,-category,-image,-languages) %>%
  rename(Date = na)

```


```{r,  echo=FALSE, message=FALSE, warning=FALSE}

library(kableExtra)

options(digits = 2, scipen = 999, big.mark = ".")

facebook %>%
  mutate(impressions = as.numeric(impressions),
         comments = as.numeric(comments),
         engagements = as.numeric(engagements),
         likes = as.numeric(likes),
         audience = as.numeric(audience)) %>%
  rename(Activity = type) -> fbDta

fbDta %>%
  group_by(author) %>%
  top_n(.,5, wt = "impressions") %>% 
  arrange(desc(impressions)) %>%
  filter(impressions > 500) %>%
  select(author, impressions, engagements, likes, text, Activity,  url, Date) %>%
  kable(format.args = list(decimal.mark = ',', big.mark = ".")) %>%
    kable_styling() %>%
    scroll_box(width = "770px", height = "450px")
```

---
# Social Media
#### Croatian influencers on Facebook (*like*)


<br>
```{r,  echo=FALSE, message=FALSE, warning=FALSE}

fbDta %>%
  top_n(.,10, wt = "likes") %>% 
  arrange(desc(likes)) %>%
  filter(likes > 50) %>%
  select(text, Activity, author, url, impressions, engagements, likes, Date) %>%
  kable(format.args = list(decimal.mark = ',', big.mark = ".")) %>%
    kable_styling() %>%
    scroll_box(width = "770px", height = "450px")

```
---
# Social Media
### Twitter
* The most famous microblogging service
<br>
<br>
* Numerous papers and analyses
<br>
<br>
* [Popular in quantitative political science](http://pablobarbera.com/static/social-media-data-generators.pdf)
<br>
<br>
* [Predicting political outcomes](https://core.ac.uk/reader/11310530)
<br>
<br>
* [Prediction of stock trends](http://cs229.stanford.edu/proj2011/GoelMittal-StockMarketPredictionUsingTwitterSentimentAnalysis.pdf)
<br>
<br>
* [Tracking public changes](https://openknowledge.worldbank.org/bitstream/handle/10986/22656/The0pulse0of0p0eform0in0El0Salvador.pdf?sequence=1&isAllowed=y)
<br>
<br>

.footnote[[*] [Support for analysis in R](https://towardsdatascience.com/a-guide-to-mining-and-analysing-tweets-with-r-2f56818fdd16); [InfoCoV](https://www.netokracija.com/infocov-istrazivanje-ana-mestrovic-covid-19-174006)]





---
# Social Media
### YouTube in Croatia (analysis at [link](https://www.netokracija.com/analiza-hrvatske-youtube-scene-od-1400-kanala-samo-njh-6-ima-preko-milijun-pretplatnika -180868))
<br>
```{r  echo=FALSE, fig.height=12}
knitr::include_url("https://www.netokracija.com/analiza-hrvatske-youtube-scene-od-1400-kanala-samo-njih-6-ima-preko-milijun-pretplatnika-180868")
```
---
# App Data
#### Primjer analize korištenja vlastitih mobilnih aplikacija 

```{r warning=FALSE}
# Učitaj pakete
library(jsonlite)
library(tidyverse)
library(patchwork)
# Učitaj podatke
android <- jsonlite::fromJSON("./02-data/MyActivity.json")
# pregledaj podatke
glimpse(android)
# prilagodi datum
android %>% 
  mutate(date = lubridate::date(time), month =lubridate::month(time))->android
```

---
# App Data
#### Example of analysis of the use of own mobile applications 
<br>
```{r}
# napravi grafikon
najAplikacije <- android %>% 
  count(header, sort = TRUE)  %>% 
  head(20) %>% 
  mutate(header = fct_reorder(header, n)) %>% 
  ggplot() + geom_col(aes(y = header, x = n)) +
  theme_minimal() +
  ylab("") +
  xlab("") +
  labs(title = "Najkorištenije aplikacije",
       subtitle = "Korištenje vlastitog Android mobilnog uredaja",
       caption = "Podatci:Google Takeout")
```

---
# App Data
#### Example of analysis of the use of own mobile applications


```{r fig.width=7, fig.height=4}
# prikaži grafički 
najAplikacije
```

---
# App Data
#### Example of analysis of the use of own mobile applications

```{r warning= FALSE, message=FALSE}
android %>% 
  filter(month %in% '7') %>% # ponovi i za listopad/nije prikazano zbog prostora
  group_by(month, header) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% #View()
  mutate(header = fct_reorder(header, n)) %>% 
  ggplot() + geom_col(aes(y = header, x = n)) +
  theme_minimal() +
  ylab("") +
  xlab("") +
  labs(title = "Najkorištenije aplikacije u srpnju",
       subtitle = "Korištenje vlastitog Android mobilnog uredaja",
       caption = "Podatci:Google Takeout") -> najAplikacije7
```

```{r echo=FALSE, warning= FALSE, message=FALSE}
android %>% 
  filter(month %in% '10') %>% 
  group_by(month, header) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% #View()
  mutate(header = fct_reorder(header, n)) %>% 
  ggplot() + geom_col(aes(y = header, x = n)) +
  theme_minimal() +
  ylab("") +
  xlab("") +
  labs(title = "Najkorištenije aplikacije u listopadu",
       subtitle = "Korištenje vlastitog Android mobilnog uredaja",
       caption = "Podatci:Google Takeout") -> najAplikacije10

```

---
# App Data
#### Example of analysis of the use of own mobile applications 

```{r  fig.width=7, fig.height=4, echo=FALSE}
# prikaži grafički
najAplikacije7 / najAplikacije10
```

---

# Sensory devices

<br>
<br>
<br>
<br>
 1. Mobile sensors (number of passengers in public transport)
<br>
<br>
 2. GPS (delivery, epidemiology)
<br>
<br>
 3. Card transactions (forecasting bankruptcy, marketing, consumer behavior)

.footnote[[*] Legally sensitive data!]


---
# Sensory devices
### Google mobility

```{r ,fig.width=6, fig.height=2}

df %>% # podatci iz paketa 
  filter(iso3c == "HRV", date >= "2020-02-21") %>%
  mutate(gov_interventions = (soc_dist + mov_rest)/
           max(soc_dist + mov_rest, na.rm = TRUE),
         lockdown = lockdown == 1) %>%
  select(date, lockdown, gov_interventions, starts_with("gcmr_")) %>%
  pivot_longer(cols = c(5:9), names_to = "Mjere", values_to = "value") %>%
  na.omit() -> dta

mobilnost <- ggplot(dta, aes(x = date, y = value, group = Mjere, color = Mjere)) + 
  theme_minimal() +
  annotate("rect", xmin = min(dta$date[dta$lockdown]), xmax = max(dta$date), 
           ymin = -Inf, ymax = Inf, 
           fill = "lightblue", color = NA, alpha = 0.2) +
  geom_line() +
  xlab("Vrijednost") +
  ylab("Datum")
```

---
# Sensory devices
### Google mobility

```{r ,fig.width=6, fig.height=3}
# prikaži grafički
mobilnost
```


---

# Big (Modern) Data Management 

<br>
<br>
<br>
.pull-left[
 **Data collection**
* API
* Web Scraping
* Databases
<br>
<br>
]

.pull-right[
**Data structuring**
* Cleaning
* Manipulation
* Programmatic approach
* Standardization
* Merge bases

<br>
<br>
]


---

# Big (Modern) Data Management 
<br>
<br>
<br>

.pull-left[
**Data analysis**
* Supervised and unsupervised machine learning
* Deep learning
* Ensemble methods
* Bayes
* NLP (txt.)
<br>
<br>
]

.pull-right[
**Robustness check**
* Information criteria
* 'Cross'-validation
* Train-test split
* Backtesting
* Accuracy of the classifier (ROC, Confusion Mtx)
<br>
<br>
]


---
class: inverse, center, middle

# Next lecture: Introduction to R.
<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>






















